---
title: "Ocean Acidification Problem Set"
author: "Juliana Olsen-Valdez & Stephanie Plaza-Torres"
date: "Last knitted `r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document: 
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options:
  chunk_output_type: inline
---

> Click into a code cell (the gray blocks below) to select or edit it. To run a selected code cell, hit `Ctrl`+`Shift`+`Enter` (Windows & Linux) / `Command`+`Shift`+`Enter` (MacOS). Make sure that each code cell runs successfully before you move on to the next one. There is also a small green 'play' button in the corner that wil run the code cell for you.

<p>&nbsp;</p>
# ***Ocean Acidification***
<p>&nbsp;</p>

##Using Data Sets 

> We'll first load some R packages that will allow us to make our code do things like pull in data sets from a folder or make a plot. *R packages* are collections of functions and data sets developed by the community. They improve the existing base fuctionalities of R and expand what you can achieve with your code. 
But what is a function?

> A *function* is a conjunction of code that achieves a specific task. This code can be 'called' using the function's given name. What the function ultimately does is take in data and return a result based on that function's 'inner' code. 

```{r load packages}
# The library() function loads each package. You need to do this so R can find the functions you need from a certain package. Once the package loads all the different functions associated with it are usable. 
library(tidyverse) #these are all different libraries that help the code do different things
library(readxl)#the blue "library" part tells R that this is a package it should load and the black writing is the name of the specific package(in this case "readxl" which helps us read Excel files into R)
library(ggplot2)#this library, for example, helps us make plots of our data. 
#notice that because there is a "#" symbol before all of the note text above, it's all just comments that don't affect the code!
```

###Modern $CO_2$ Concentration

At Mauna Loa, NOAA has data looking at the parts per million concentration of $CO_2$


Now, we'll load PETM data of estimated SST (degrees Celsius):
```{r load data}
PETM_data <- #we are making a new data set called "PETM_data" 
  read_excel("data/FAKE_PETM_SST_data.xlsx") #we are reading in the excel file with our data (which is in the data folder)
```

Plot the estimated SST (degrees Celcius) data over time:
```{r plot PETM data}
SSTvAge <- #We create a new plot that's named "SSTvAge"
ggplot(PETM_data, aes(x = Age, y = SST)) + #we call the function ggplot to input our data set and assign x and y variables
geom_line(size = 0.5, color = 'red')+ #We define a red line to connect our data points
labs(y = "SST (degrees C)", x = "Age (Ma)", title = "SST (degrees Celcius) vs Age (Ma) during the PETM from Stokke et al. (2020)" )+ #We give our graph labels
theme_bw() #We make the background grid theme/color black and white 
SSTvAge
```

##Sources: 
- McInerney, F. A., & Wing, S. L. (2011). The Paleocene-Eocene Thermal Maximum: a perturbation of carbon cycle, climate, and biosphere with implications for the future. Annual Review of Earth and Planetary Sciences, 39, 489-516.
- NOAA Global Monitoring Laboratory, Mauna Loa
- NOAA Extended Reconstructed Sea Surface Temperature (ERSST)
- PETM data modified from Stokke, E. W., Jones, M. T., Tierney, J. E., Svensen, H. H., & Whiteside, J. H. (2020). Temperature changes across the Paleocene-Eocene Thermal Maximum–a new high-resolution TEX86 temperature record from the Eastern North Sea Basin. Earth and Planetary Science Letters, 544, 116388.



